---
title: "Merge data"
author: "Duc-Quang Nguyen | swissinfo.ch"
date: " 2016"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: simplex
---

# Data

* Depuis 2010: [Permanent and non permanent resident population by canton, sex, residence permit, age class and citizenship](https://www.pxweb.bfs.admin.ch/Selection.aspx?px_language=en&px_db=px-x-0103010000_101&px_tableid=px-x-0103010000_101%5Cpx-x-0103010000_101.px&px_type=PX) [Complete px cube](https://www.pxweb.bfs.admin.ch/DownloadFile.aspx?file=px-x-0103010000_101)
  * the whole dataset is huge. STAT-TAB query
      * Select all years
      * Select only Switzerland
      * Population type BOTH/ALL
      * Sexe NULL
      * Autorisation de résidence NULL
      * Classe d'âge NULL
      * Nationalité select ALL
  * File saved as comma delimited [px-x-0103010000_101.csv](data/px-x-0103010000_101.csv)

* 1850-2009: [Population résidante permanente et non permanente selon le sexe et la nationalité par pays](https://www.bfs.admin.ch/bfs/fr/home/statistiques/catalogues-banques-donnees.assetdetail.hs-f-01.01.01.03.html)
   * **Warning: Turkey is considered a European country**

```{r setup, include=FALSE}
output_base.file <- "input/merged_foreigners"
outputSwissFor_base.file <- "input/merged_Swisssforeigners"

testPlot <- T

natpop.file1 <- "data/hs-e-01.01.01.03.xls" #data/hs-f-01.01.01.03.csv"
natpop.file2 <- "data/px-x-0103010000_101.csv"

# cluster ex-yougoslavia countries
Yugoslavia <- c('Yugoslavia', 'Serbia and Montenegro', 'Serbia', 'Montenegro', 
                'Kosovo', 'Croatia', 'Bosnia and Herzegovina', 'Macedonia')
# cluster NA nationalities
Unknown <- c('No indication', 'Stateless', 'Not attributable according to current borders')


library(readr)
library(readxl)
library(tidyr)
library(dplyr)
library(magrittr)
library(countrycode)
library(ggplot2)
library(scales)
library(swiMap)
library(swiTheme)
```

## Load historical data

```{r shape 1850-2009 xls data}
data.read <- read_excel(natpop.file1, skip  = 2)

# drop gender rows
data.read <- data.read[!unlist(data.read[,1]) %in% c("Male", "Female"),]
# drop footer
data.read <- data.read[!is.na(unlist(data.read[,1])),]

# drop trailing number), i.e. 3)
rown <- gsub("\\d\\)", "", unlist(data.read[,1]))
# drop trailing space
rown <- gsub(" $", "", rown)

# rename country/regions
rown <- gsub("Stateless/No indication", "Unknown", rown)
rown <- gsub("Austalia/Oceania", "Oceania", rown)

# remove 3 dots by NA
data.read <- data.read[,-c(1)]
d1 <- apply(data.read, 2, function(l) as.numeric(gsub("...", "", l, fixed = T))) %>% as.data.frame()

colnames(d1) <- as.numeric(colnames(d1))
d1 <- cbind(nationality = rown, d1)

## make the data long
d1 <- tidyr::gather(d1, year, value, -nationality)
d1$year <- as.numeric(d1$year)

## checks
# Total == Swiss + Foreigner
check.diff <- d1 %>% filter(nationality == "Total") %>% select(value) %>% unlist() - 
  ((d1 %>% filter(nationality == "Swiss") %>% select(value) %>% unlist()) + 
     (d1 %>% filter(nationality == "Foreigner") %>% select(value) %>% unlist())
  )

if(!all(check.diff == 0)) {
  warning("Sum of foreigners + Swiss is always equal to Total")
  print(check.diff)
}

nat.group <- c('Total', 'Swiss', 'EU/EFTA country', 'Europe', 'Foreigner')

# Foreigner = everything (except Total,  Swiss, 'EU/EFTA country', 'Europe')
check.diff <- d1 %>% filter(nationality == 'Foreigner') %>% select(value) %>% unlist() - 
  (d1 %>% filter(!nationality %in% nat.group) %>% 
     group_by(year) %>% summarise(value = sum(value, na.rm = T)) %>% 
     ungroup() %>% select(value) %>% unlist())
if(!all(check.diff == 0)) {
  warning("Sum of nationality (- some) not equal to Foreigner")
  print(check.diff)
}

## Merge & drop nationalities
#levels(d1$nationality)

d1.g1 <- d1 %>% filter(nationality %in% nat.group)
# discard groups
d1 <- d1 %>% filter(!nationality %in% nat.group)

# Merge ex-Yougoslavia nationalities
yugo <- d1 %>% filter(nationality %in% Yugoslavia) %>%
  group_by(year) %>% summarise(value = sum(value, na.rm = T)) %>%
  ungroup()
d1 <- rbind(
  d1 %>% filter(!nationality %in% Yugoslavia),
  cbind(nationality = 'Ex-Yugoslavia', yugo)
) %>% arrange(year)

# rename America to Americas (to be consistent with countrycode continent)
d1$nationality <- as.character(d1$nationality)
d1[which(d1$nationality == "America"), 'nationality'] <- "Americas"
#d1$nationality<- factor(d1$nationality)

# check
chkb <- d1 %>% group_by(year) %>% summarise(value = sum(value, na.rm = T)) %>%
  select(value) %>% unlist()
chka <- d1.g1 %>% filter(nationality == "Foreigner") %>% select(value) %>% unlist()

stopifnot((chka - chkb) == check.diff)
```

## Data from 2010

```{r wrangle data from 2010 onwards}
data.read <- read_tsv(natpop.file2, skip = 2, locale = locale(encoding = "latin1"))

# drop & merge pop type
data.read %<>% select(-Canton, -`Population type`) %>%
  rename(Total = `Citizenship - Total`, Swiss = Switzerland, year = Year)
# make data long
d2 <- data.read %>% gather(nationality, value,-year) %>%
  group_by(year, nationality) %>% summarise(value = sum(value, na.rm = T)) %>%
  ungroup() %>% arrange(year, desc(value))
# remove NA rown

# filter out Swiss and Total numbers
d2.g2 <- d2 %>% filter(nationality %in% c('Total', 'Swiss'))
d2 %<>% filter(!nationality %in% c('Total', 'Swiss'))

d2.g2 <- rbind(d2.g2,
          cbind(
            nationality = "Foreigner", 
            d2 %>% group_by(year) %>% summarise(value = sum(value, na.rm = T)) %>% ungroup()
          ) 
      ) %>% arrange(year)

# check total is a sum of all other nationality
check.diff <- d2.g2 %>% filter(nationality == "Total") %>% select(value) %>% unlist() -
  (
    d2.g2 %>% filter(nationality != "Total") %>% group_by(year) %>%
      summarise(value = sum(value)) %>% select(value) %>% ungroup() %>% unlist()
  )
stopifnot(check.diff == 0)

## Create similar groups to d1
fnat <- as.character(unique(d1$nationality))

d2$iso2 <- countrycode(d2$nationality, "country.name", "iso2c")
d2$continent <- countrycode(d2$nationality, "country.name", "continent")
# hack fix for Taiwan
d2[which(d2$iso2 == "TW"), 'continent'] <- 'Asia'

d2.bak <- d2

d2$nat <- d2$nationality
# relabel nationality for Ex-Yugoslavia and Unknown
idx <- which(d2$nationality %in% Yugoslavia | d2$iso2 %in% na.omit(countrycode(Yugoslavia, "country.name", "iso2c")))
cat("\n", d2[idx,'nat'] %>% unlist(use.names = F) %>% unique(), "\t these countries about to be merged!\n")
d2[idx,'nat'] <- "Ex-Yugoslavia"
idx <- which(d2$nationality %in% Unknown | (is.na(d2$iso2) & d2$nat != "Ex-Yugoslavia"))
cat("\n", d2[idx,'nat'] %>% unlist(use.names = F) %>% unique(), "\t these countries about to be merged!\n")
d2[idx, 'nat'] <- "Unknown"

# get the country if in subset country otherwise get the continent
nat <- ifelse(d2$nat %in% fnat, d2$nat, d2$continent)

# rename Europe to "Other European country" to be consistent with the historical data
nat <- ifelse(nat == "Europe", "Other European country", nat)

stopifnot(fnat %in% nat)
stopifnot(!any(is.na(nat)))
stopifnot(length(nat) == nrow(d2))

# recompute value by nat
d2$nat <- nat
d2 %<>% group_by(nat, year) %>%
  summarise(value = sum(value)) %>% 
  ungroup() %>% rename(nationality = nat)

dd <- bind_rows(d1, d2)
outfile <-  paste0(output_base.file, "_", paste0(range(dd$year), collapse = "_"), ".csv")
cat("\n", "The merged timeseries to be written at: ", outfile)
write.csv(dd, file = outfile, row.names = F)
```

```{r save the merged timeseries for Swiss & total foreigners}
d1.g1 %<>% filter(nationality %in% c('Foreigner', 'Swiss'))
d2.g2 %<>% filter(nationality %in% c('Foreigner', 'Swiss'))

dg <- suppressWarnings (bind_rows(d1.g1, d2.g2 %>% select(nationality, year, value)))
dg %<>% spread(nationality, value)
#dg$Foreigner / (dg$Swiss + dg$Foreigner)

outfile2 <-  paste0(outputSwissFor_base.file, "_", paste0(range(dd$year), collapse = "_"), ".csv")

cat("\n", "The merged timeseries of Swiss and foreigners to be written at: ", outfile2)
write.csv(dg, file = outfile2, row.names = F)
```


```{r test plot, echo = F}
if(testPlot) {
  ggplot(data = dd) + geom_area(aes(year, value, group = nationality, fill = nationality)) + swi_theme()
  
  ggplot(data = dd) + geom_area(aes(year, value, group = nationality, fill = nationality)) + facet_wrap(~nationality) + swi_theme()
  

}
```